NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2019-02-05 Manuscript/manuscriptPS.tex
Stylistics and trying to improve clarity.

\author{Pavel Buslaev}
\affiliation{Moscow Institute of Physics and Technology}
\author{Fernando Favela}
\affiliation{Mexico}
\author{Tiago M. Ferreira}
\affiliation{Halle, Germany}
\author{Ivan Gushchin}
\affiliation{Moscow Institute of Physics and Technology}
\author{Matti Javanainen}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{Batuhan Kav}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Jesper J. Madsen}
\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois 60637, United States of America}
\author{Markus Miettinen}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Josef Melcr}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{Ricky Nencini}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{O. H. Samuli Ollila}
\email[]{samuli.ollila@helsinki.fi}
%\homepage[]{Your web page}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\affiliation{Institute of Biotechnology, University of Helsinki}
\author{Thomas Piggot   \todo{Authorlist is not yet complete}}
\affiliation{Southampton, United Kingdom}
%\author{NMRlipids collaboration}
%\affiliation{nmrlipids.blogspot.fi}

@@

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.5cm]{../Figs/fig1_POPS.pdf}
  \caption{\label{PShgSIGNSsimpson}
    The headgroup and glycerol backbone region of the (A) INEPT spectrum and
    (B) 2D R-PDLF spectra.
    (C) Experimental SDROSS data (points) and SIMPSON simulations (blue lines) with
    the order parameter values of -0.12 for the $\beta$-carbon, and 0.09 and -0.02
    for the $\alpha$-carbon slittings. The S-DROSS curve from SIMPSON simulation with positive value
    for the smaller $\alpha$-carbon order parameter (dashed grey).
  }
  \todo{I think that the peak labeling would be good to show also in (A).}
\end{figure}
\begin{figure}[!htb]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PCPScomp-eps-converted-to.pdf}
  \caption{\label{HGorderParameters}
    (A) Chemical structures and labels for the headgroup and glycerol backbone carbons.
    (B) Headgroup and glycerol backbone order parameters of POPS measured in this work compared
    with the previously published values from DOPS ($^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
    POPC  ($^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
    are measured in this work. Signs of the PC order parameters are measured previously~\cite{ferreira16}.
  }
\end{figure}

@@

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.pdf}
  \caption{\label{HGorderParametersPS}
    Order parameters of PS headgroup ($\beta$ and $\alpha$) and
    glycerol backbone (g$_3$, g$_2$, g$_1$) from NMR experiments (horizontal lines),
    and MD simulations with different force fields (dots).
    Experimental data for DOPS are measured with 0.1~M of NaCl~\cite{browning80},
    while all the other data are without additional salt.
    All the data for DOPS at 303~K and all the data for POPS at 298~K.
    Light blue areas span 0.04 units around the average of the extremal experimental values,
    in accordance with the expected quantitative accuracy of experimental values~\cite{ollila16}.
    The vertical bars shown for all simulation values except MacRog~K$^+$
    are not error bars, but demonstrate that for these systems
    we had at least two data sets; the ends of the bars mark the extreme values
    from the sets, and the dot marks their measurement-time-weighted average.
  }
\end{figure}

@@

The headgroup and glycerol backbone of PS lipids show a wide variety in the order parameters (Fig.~\ref{HGorderParametersPS})
and structures (Fig.~\ref{HGandGLYstructuresPS}) between different simulation models,
as previously observed also for PC lipids \cite{botan15}.
The models perform generally less well for PS than for PC
(Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs. 2 and 4 in Ref.~\cite{botan15}).
Therefore, the interpretation of structural differences between PC and PS headgroups
from simulations is not straightforward.

@@

However, if one excludes the glycerol backbone and focuses only on the headgroup ($\alpha$ and $\beta$) order parameters,
the best performing models,
Slipids, CHARMM36 and CHARMM36ua, do reproduce the larger-than-in-PC forking of the $\alpha$-carbon.
Slipids reproduces also the feature that the $\beta$-carbon order parameter
in PS is significantly smaller than in PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15}).
Interestingly, the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$  dihedral with
a single and narrow peak in the angle distribution close to 120$^{\circ}$ is more restricted
in the best three models than in other models which give two maxima with different angles (Fig. \ref{dihedralsHG}).
The restricted motion is also
visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} (A) and \ref{HGandGLYstructuresPS})
suggesting that the rotation of the carboxyl group is limited in the serine headgroup.
In addition, the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibits a more asymmetric
and restricted angle distribution for PS than for PC headgroup in
CHARMM36 simulations in best agreement with experiments
(Figs.~\ref{HGstructuresPSandPC} (B) and \ref{dihedralsHGpc}).
The results might manifest the increased rigidity anticipated in the early experimental studies \cite{browning80,buldt81}.
Also, the sampled conformations of glycerol backbone
significantly vary between different simulation models 
(Figs. \ref{dihedralsGLY}, \ref{HGandGLYstructuresPS}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}),
but further analysis is beyond the scope of this work which focuses on the PS headgroup.

The suggested characteristic conformations of the PS headgroup may be useful when interpreting experiments and
can guide further force field development. However, more accurate MD force fields are
required to confirm the suggested conformations because the simulated PS headgroup order parameters
are not within the experimental error in any of the tested models.

@@

Calcium binding affinity to membranes containing negative charged PS lipids can be
measured by detecteding the PC lipid headgroup order parameters
from POPC:POPS (5:1) mixtures (section~\ref{electrometerFORmixtures}),
where the dehydrated lipid-ion complexes and phase separation are 
not observed~\cite{feigenson86,mattai89,roux90,roux91}.
Despite the lack of ion free reference state
in the presence of negatively charged lipids, the simulations give
coherent results for POPC headgroup order parameters as a function of
CaCl$_2$ in the POPC:POPS (5:1) mixtures (Fig. \ref{changesWITHCaClPS}).
As expected from the previous study of pure PC lipid
bilayers \cite{catte16}, almost all the tested simulation models overestimate the
experimentally observed~\cite{roux90} decrease of the POPC headgroup order parameters
in POPC:POPS (5:1) mixtures as a function of Ca$^{2+}$ concentration (Fig. \ref{changesWITHCaClPS}),
indicating overestimated calcium binding binding affinity.
Only exception is the CHARMM36 model with the NBfix
interaction employed for calcium~\cite{kim16}, which underestimates the changes in order parameter 
indicating weaker binding affinity than experiments.
%incorporated in the parameters give by the CHARMM-GUI at the time of running the simulations (January 2018),
Notably, CHARMM36 simulations with the NBfix corrections \cite{venable13,kim16} give similar binding affinities of
calcium and sodium to POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to the experimental 
data~\cite{cevc90,akutsu81,altenbach84}. Therefore, we conclude that the calcium binding affinity
is underestimated in CHARMM36 simulations with the NBfix for calcium \cite{kim16} but overestimated 
in all the other tested models. This is evident in the calcium density distributions
along membrane normal where almost all Ca$^{2+}$ ions bind to the membrane interface in
all simulation models except CHARMM36 (Fig. \ref{CAdensPCPSmixture}).

The headgroup order parameters of POPS measured from POPC:POPS (5:1) mixture
exhibit a strong dependence of CaCl$_2$ with small concentrations and rapid saturation
below 100 mM (Fig. \ref{changesWITHCaClPS}).
The order parameter of POPS $\beta$-carbon increase and the larger $\alpha$-carbon 
decrease with the added CaCl$_2$ in experiments. Slight increase is observed in
the smaller $\alpha$-carbon. All the changes are significantly overestimated in the
tested simulation models, including the CHARMM36 with underestimated binding affinity.
In addition, different simulation models predict qualitatively different behaviour
for the POPS $\alpha$-carbon order parameters with the added calcium.
For example, both order parameters decrease in Berger simulations but increase
in MacRog simulations, while behaviour in Lipid14/17 and CHARMM36 simulations is
more complicated. This is in contrast to the PC headgroup, where
qualitatively correct reponse to the bound ions is observed
in all simulation models despite of the significant discrepancies in the headgroup
structure without additional ions~\cite{catte16}. Therefore, we conclude that
the improvement of force fields is necessary to correctly describe interactions between
PS headgroup and calcium ions in MD simulations.

>>>>

\author{Pavel Buslaev}
\affiliation{Moscow Institute of Physics and Technology}
\author{Fernando Favela}
\affiliation{Mexico}
\author{Tiago M. Ferreira}
\affiliation{Halle, Germany}
\author{Ivan Gushchin}
\affiliation{Moscow Institute of Physics and Technology}
\author{Matti Javanainen}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{Batuhan Kav}
\affiliation{Potsdam, Germany}
\author{Jesper J. Madsen}
\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois 60637, United States of America}
\author{Markus Miettinen}
\affiliation{Potsdam, Germany}
\author{Josef Melcr}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{Ricky Nencini}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{O. H. Samuli Ollila}
\email[]{samuli.ollila@helsinki.fi}
%\homepage[]{Your web page}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\affiliation{Institute of Biotechnology, University of Helsinki}
\author{Thomas Piggot   \todo{Authorlist is not yet complete}}
\affiliation{Southampton, United Kingdom}
%\author{NMRlipids collaboration}
%\affiliation{nmrlipids.blogspot.fi}

@@

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.5cm]{../Figs/fig1_POPS.pdf}
  \caption{\label{PShgSIGNSsimpson}
    The headgroup and glycerol backbone region of the (A) INEPT spectrum and
    (B) 2D R-PDLF spectra.
    (C) Experimental SDROSS data (points) and SIMPSON simulations (blue lines) with
    the order parameter values of -0.12 for the $\beta$-carbon, and 0.09 and -0.02
    for the $\alpha$-carbon slittings. The S-DROSS curve from SIMPSON simulation with positive value
    for the smaller $\alpha$-carbon order parameter (dashed grey).
  }
  \todo{I think that the peak labeling would be good to show also in (A).}
\end{figure}
\begin{figure}[!htb]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PCPScomp.eps}
  \caption{\label{HGorderParameters}
    (A) Chemical structures and labels for the headgroup and glycerol backbone carbons.
    (B) Headgroup and glycerol backbone order parameters of POPS measured in this work compared
    with the previously published values from DOPS ($^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
    POPC  ($^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
    are measured in this work. Signs of the PC order parameters are measured previously~\cite{ferreira16}.
  }
\end{figure}

@@

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.pdf}
  \caption{\label{HGorderParametersPS}
    Order parameters of PS headgroup and glycerol backbone from experiments
    and simulations with different force fields.
    Experimental data for DOPS is measured with 0.1~M of NaCl~\cite{browning80},
    while all the other data is without additional ions.
    All the data for DOPS is at 303~K and all the data for POPS is at 298~K.
    Light blue areas span 0.04 units around the average of the extremal experimental values,
    which is the expected quantitative accuracy of experimental values~\cite{ollila16}.
    The vertical bars shown for some simulation values
    are not error bars, but demonstrate that for these systems
    we had at least two data sets; the ends of the bars mark the extreme values
    from the sets, and the dot marks their measurement-time-weighted average.
  }
\end{figure}

@@

The headgroup and glycerol backbone of PS lipids show wide variety in the order parameters and structures
between different simulation models (Figs. \ref{HGorderParametersPS} and \ref{HGandGLYstructuresPS}),
as previously observed also for PC lipids \cite{botan15}.
The models perform generally less well for PS lipids than for PC lipids 
(Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs. 2 and 4 in Ref.~\cite{botan15}).
Therefore, the interpretation of structural differences between PC and PS headgroups
from simulations is not straightforward.

@@

However, the best performing models,
Slipids, CHARMM36 and CHARMM36ua, reproduce the larger forking of the $\alpha$-carbon 
and the Slipids model reproduces also the lower of the $\beta$-carbon order parameter
when comparing the PS results to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15}).
Interestingly, the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$  dihedral with
a single and narrow peak in the angle distribution close to 120$^{\circ}$ is more restricted
in the best three models than in other models which give two maxima with different angles (Fig. \ref{dihedralsHG}).
The restricted motion is also
visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} (A) and \ref{HGandGLYstructuresPS})
suggesting that the rotation of carboxyl group is limited in the serine headgroup.
In addition, the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibit a more asymmetric
and restricted angle distribution for PS than for PC headgroup in
CHARMM36 simulations in best agreement with experiments
(Figs.~\ref{HGstructuresPSandPC} (B) and \ref{dihedralsHGpc}).
The results might manifest the increased rigidity anticipated in the early experimental studies \cite{browning80,buldt81}.
Also, the sampled conformations of glycerol backbone
significantly vary between different simulation models 
(Figs. \ref{dihedralsGLY}, \ref{HGandGLYstructuresPS}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}),
but further analysis is beyond the scope of this work which the PS headgroup.

The suggested characteristic conformations of PS headgroup may be useful when interpreting experiments and
can guide the further force field development. However, more accurate simulations are
required to confirm the suggested conformations because the simulated PS headgroup order parameters
are not within the experimental error bars in any of the tested models.

@@

Calcium binding affinity to membranes containing the negative charged PS lipids can be
experimentally measured by detecting the PC lipid headgroup order parameters
from POPC:POPS (5:1) mixtures (section~\ref{electrometerFORmixtures}),
where the dehydrated lipid-ion complexes and phase separation are 
not observed~\cite{feigenson86,mattai89,roux90,roux91}.
Despite the lack of an ion-free reference state
in the presence of negatively charged lipids, our simulations give
coherent results for POPC headgroup order parameters as a function of
CaCl$_2$ in the POPC:POPS (5:1) mixtures (Fig. \ref{changesWITHCaClPS}).
As expected from the previous study of pure PC lipid
bilayers \cite{catte16}, almost all the tested simulation models overestimate the
experimentally observed~\cite{roux90} decrease of the POPC headgroup order parameters
in POPC:POPS (5:1) mixtures as a function of Ca$^{2+}$ concentration (Fig. \ref{changesWITHCaClPS}),
indicating overestimated calcium binding affinity.
The only exception is the CHARMM36 model with the NBfix
interaction employed for calcium~\cite{kim16}, which underestimates the changes in order parameters, 
indicating weaker binding affinity than experiments.
%incorporated in the parameters give by the CHARMM-GUI at the time of running the simulations (January 2018),
Notably, CHARMM36 simulations with the NBfix corrections \cite{venable13,kim16} give similar binding affinities of
calcium and sodium to POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to the experimental 
data~\cite{cevc90,akutsu81,altenbach84}. Therefore, we conclude that the calcium binding affinity
is underestimated in CHARMM36 simulations with the NBfix for calcium \cite{kim16}, but overestimated 
in all the other tested models. This is evident in the calcium density distributions
along membrane normal, where almost all Ca$^{2+}$ ions bind to the membrane interface in
all simulation models except CHARMM36 (Fig. \ref{CAdensPCPSmixture}).

The headgroup order parameters of POPS experimentally measured from a POPC:POPS (5:1) mixture
exhibit a strong dependence of CaCl$_2$ with small concentrations and rapid saturation
below 100 mM (Fig. \ref{changesWITHCaClPS}).
In experiments, the order parameter of the POPS $\beta$-carbon increases with added CaCl$_2$,
whereas the larger $\alpha$-carbon order parameter decreases; a slight increase is observed in
the smaller $\alpha$-carbon. All these changes are significantly overestimated in the
tested simulation models, including CHARMM36 with underestimated binding affinity.
In addition, different simulation models predict qualitatively different behaviour
for the POPS $\alpha$-carbon order parameters with added calcium.
For example, both order parameters decrease in Berger, but increase
in MacRog, and in Lipid14/17 and CHARMM36 a more complicated behavior is seen.
This is in contrast to the PC headgroup, where
qualitatively correct reponse to bound ions is observed
in all simulation models, despite significant discrepancies in the headgroup
structure without additional ions~\cite{catte16}. Therefore, we conclude that
improvement of force fields is necessary to correctly describe interactions between the
PS headgroup and calcium ions in MD simulations.
